---
title: "hal90001 Benchmarks"
author: "Jeremy Coyle"
date: "10/5/2017"
output: html_document
---

```{r setup, include=FALSE, results='hide'}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sl3)
library(delayed)
library(SuperLearner)
library(future)
library(ggplot2)
library(data.table)
library(stringr)
library(scales)
```

## Introduction

This document consists of some simple benchmarks for various choices of SuperLearner implementation, wrapper functions, and parallelization schemes. The purpose of this document is two-fold: 

1. Compare the computational performance of these methods
2. Illustrate the use of these different methods

## Test Setup


### Test System

```{r systemInfo, echo=FALSE, results="asis"}
uname <- system("uname -a", intern = TRUE)
os <- sub(" .*", "", uname)
if(os=="Darwin"){
  cpu_model <- system("sysctl -n machdep.cpu.brand_string", intern = TRUE)
  cpus_physical <- as.numeric(system("sysctl -n hw.physicalcpu", intern = TRUE))
  cpus_logical <- as.numeric(system("sysctl -n hw.logicalcpu", intern = TRUE))
  cpu_clock <- system("sysctl -n hw.cpufrequency_max", intern = TRUE)
  memory <- system("sysctl -n hw.memsize", intern = TRUE)
} else if(os=="Linux"){
  cpu_model <- system("lscpu | grep 'Model name'", intern = TRUE)
  cpu_model <- gsub("Model name:[[:blank:]]*","", cpu_model)
  cpus_logical <- system("lscpu | grep '^CPU(s)'", intern = TRUE)
  cpus_logical <- as.numeric(gsub("^.*:[[:blank:]]*","", cpus_logical))
  tpc <- system("lscpu | grep '^Thread(s) per core'", intern = TRUE)
  tpc <- as.numeric(gsub("^.*:[[:blank:]]*","", tpc))
  cpus_physical <- cpus_logical/tpc
  cpu_clock <- as.numeric(gsub("GHz","",gsub("^.*@","",cpu_model)))*10^9
  memory <- system("cat /proc/meminfo | grep '^MemTotal'", intern = TRUE)
  memory <- as.numeric(gsub("kB","",gsub("^.*:","",memory)))*2^10
} else {
  stop("unsupported OS")
}
```

* CPU model: `r cpu_model`
* Physical cores: `r as.numeric(cpus_physical)`
* Logical cores: `r as.numeric(cpus_logical)`
* Clock speed: `r as.numeric(cpu_clock)/10^9`GHz
* Memory: `r round(as.numeric(memory)/2^30, 1)`GB

### Test Data

### Tests

```{r lassi}
microbenchmark({
   glmnet::glmnet(x = x_basis, y = y, intercept = TRUE, nlambda = 100,
   lambda.min.ratio = 0.01, family = "gaussian", alpha = 1, standardize = TRUE)
 }, times = 10)
microbenchmark({
   lassi(x_basis, y, nlambda=100, lambda_min_ratio = 0.01, center = FALSE)
 }, times = 10)

microbenchmark({
   lassi(x_basis, y, nlambda=100, lambda_min_ratio = 0.01, center = TRUE)
 }, times = 10)

microbenchmark({
   glmnet::cv.glmnet(x = x_basis, y = y, intercept = TRUE, nlambda = 100,
   lambda.min.ratio = 0.01, family = "gaussian", alpha = 1, standardize = TRUE)
 }, times = 1)

microbenchmark({
   cv_lasso(x_basis, y, center = FALSE)
 }, times = 1)

microbenchmark({
   cv_lasso(x_basis, y, center = TRUE)
 }, times = 1)

set.seed(1234)
cv_l_full <- cv_lasso(x_basis, y, center = FALSE)
set.seed(1234)
cv_l_es <- cv_lasso_early_stopping(x_basis, y)
plot(cv_l_es)
plot(cv_l_full$lambdas_cvmse)
microbenchmark({
   cv_lasso_early_stopping(x_basis, y)
 }, times = 1)

```

## Session Information

```{r sessionInfo, echo=FALSE, results="asis"}
sessionInfo()
```