---
output:
  rmarkdown::github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`hal9001`

[![Travis-CI Build Status](https://travis-ci.org/tlverse/hal9001.svg?branch=master)](https://travis-ci.org/tlverse/hal9001)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/tlverse/hal9001?branch=master&svg=true)](https://ci.appveyor.com/project/tlverse/hal9001)
[![Coverage Status](https://img.shields.io/codecov/c/github/tlverse/hal9001/master.svg)](https://codecov.io/github/tlverse/hal9001?branch=master)
[![CRAN](http://www.r-pkg.org/badges/version/hal9001)](http://www.r-pkg.org/pkg/hal9001)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/hal9001)](https://CRAN.R-project.org/package=hal9001)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3558313.svg)](https://doi.org/10.5281/zenodo.3558313)

> The _Scalable_ Highly Adaptive Lasso

__Authors:__ [Jeremy Coyle](https://github.com/tlverse), [Nima
Hejazi](https://nimahejazi.org), and [Mark van der
Laan](https://vanderlaan-lab.org/)

---

## What's `hal9001`?

`hal9001` is an R package providing an implementation of the scalable _highly
adaptive lasso_ (HAL), a nonparametric regression estimator that applies
L1-regularized lasso regression to a design matrix composed of indicator
functions corresponding to the support of the functional over a set of
covariates and interactions thereof. HAL regression allows for arbitrarily
complex functional forms to be estimated at fast (near-parametric) convergence
rates under only global smoothness assumptions [@vdl2017generally;
@bibaut2019fast]. For detailed theoretical discussions of the highly adaptive
lasso estimator, consider consulting, for example, @vdl2017generally,
@vdl2017finite, and @vdl2017uniform. For a computational demonstration of the
versatility of HAL regression, see @benkeser2016hal. Recent theoretical works
have demonstrated success in building efficient estimators of complex
parameters when particular variations of HAL regression are used to estimate
nuisance parameters [e.g., @vdl2019efficient; @ertefaie2020nonparametric].

---

## Installation

For standard use, we recommend installing the package from
[CRAN](https://CRAN.R-project.org/package=hal9001) via

```{r cran-installation, eval = FALSE}
install.packages("hal9001")
```

To contribute, install the _development version_ of `hal9001` from GitHub via
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval = FALSE}
remotes::install_github("tlverse/hal9001")
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/tlverse/hal9001/issues).

---

## Example

Consider the following minimal example in using `hal9001` to generate
predictions via Highly Adaptive Lasso regression:

```{r example}
# load the package and set a seed
library(hal9001)
set.seed(385971)

# simulate data
n <- 100
p <- 3
x <- matrix(rnorm(n * p), n, p)
y <- x[, 1] * sin(x[, 2]) + rnorm(n, mean = 0, sd = 0.2)

# fit the HAL regression
hal_fit <- fit_hal(X = x, Y = y)
hal_fit$times

# training sample prediction
preds <- predict(hal_fit, new_data = x)
mean(hal_mse <- (preds - y)^2)
```

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/tlverse/hal9001/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `hal9001` R package, please cite the following:

        @manual{coyle2020hal9001,
          author = {Coyle, Jeremy R and Hejazi, Nima S and {van der Laan}, Mark
            J},
          title = {{hal9001}: The scalable highly adaptive lasso},
          year  = {2020},
          howpublished = {\url{https://github.com/tlverse/hal9001}},
          note = {{R} package version 0.2.6},
          url = {https://doi.org/10.5281/zenodo.3558313},
          doi = {10.5281/zenodo.3558313}
        }

---

## License

&copy; 2017-2020 [Jeremy R. Coyle](https://github.com/tlverse) & [Nima S.
Hejazi](https://nimahejazi.org)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

---

## References

