---
output: rmarkdown::html_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`hal9001`

[![Travis-CI Build Status](https://travis-ci.org/jeremyrcoyle/hal9001.svg?branch=master)](https://travis-ci.org/jeremyrcoyle/hal9001)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/jeremyrcoyle/hal9001?branch=master&svg=true)](https://ci.appveyor.com/project/jeremyrcoyle/hal9001)
[![Coverage Status](https://img.shields.io/codecov/c/github/jeremyrcoyle/hal9001/master.svg)](https://codecov.io/github/jeremyrcoyle/hal9001?branch=master)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)

> _Fast_ and _scalable_ estimation procedure for the Highly Adaptive LASSO

__Authors:__ [Jeremy Coyle](https://github.com/jeremyrcoyle) and [Nima
Hejazi](https://nimahejazi.org)

---

## What's `hal9001`?

`hal9001` is an R package providing an implementation of the Highly Adaptive
LASSO (HAL), a nonparametric regression estimator that applies L1-regularized
regression (i.e., the LASSO) to a design matrix composed of indicator functions
corresponding to a set of covariates and interactions thereof, in a standard
statistical learning problem. Recent theoretical results show that HAL is
endowed with several important properties that make it optimally suited for
the purpose of inference in problem settings where causal parameters are
estimated via data-adaptive techniques (i.e., machine learning), as is the case
within the framework of Targeted Minimum Loss-Based Estimation (TMLE). While it
is certainly possible to implement HAL purely in R, the computationally
intensive nature of the algorithm suggests that writing core routines in C++
(and making these available in R via the [Rcpp](http://www.rcpp.org/) framework)
ought to provide significant efficiency gains. `hal9001` is just such an
implementation.

---

## Installation

<!--
For standard use, we recommend installing the package from
[CRAN](https://cran.r-project.org/) via

```{r cran-installation, eval = FALSE}
install.packages("hal9001")
```
-->

You can install the development version of `hal9001` from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/) with

```{r gh-installation, eval = FALSE}
devtools::install_github("jeremyrcoyle/hal9001")
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/jeremyrcoyle/hal9001/issues).

---

## Example

This minimal example shows how to use `hal9001` to obtain predictions based on
the Highly Adaptive LASSO. For details on the properties of the estimator, the
interested reader is referred to @benkeser2016hal and @vdl2017generally.

```{r example}
# load the hal9001 package
library(hal9001)

# simulate data
set.seed(385971)
n = 100
p = 3
x <- xmat <- matrix(rnorm(n * p), n, p)
y <- sin(x[, 1]) * sin(x[, 2]) + rnorm(n, 0.2)

# fit the HAL regression
hal_fit <- fit_hal(X = x, Y = y)
hal_fit$times

# training sample prediction
preds <- predict(hal_fit, new_data = x)
mean(hal_mse <- (preds - y)^2)
```

---

## Contributions

`hal9001` is the primary implementation of the Highly Adaptive LASSO, an
estimator with numerous optimality properties, especially useful when employing
[Targeted Learning](http://www.targetedlearningbook.com/about/). To that end,
contributions are very welcome, though we ask that interested contributors
consult our [`contribution
guidelines`](https://github.com/jeremyrcoyle/hal9001/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `hal9001` R package, please cite the following:

        @misc{coyle2017hal9001,
          author = {Coyle, Jeremy R and Hejazi, Nima S},
          title = {{hal9001}: A fast and scalable {Highly Adaptive LASSO}},
          year  = {2017},
          howpublished = {\url{https://github.com/jeremyrcoyle/hal9001}},
          url = {http://dx.doi.org/DOI_TBD},
          doi = {DOI_TBD}
        }

---

## License

&copy; 2017 [Jeremy R. Coyle](https://github.com/jeremyrcoyle) & [Nima S.
Hejazi](http://nimahejazi.org)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

