---
output: github_document
bibliography: README-refs.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`hal9001`

> A _fast_ Highly Adaptive LASSO estimator

__Authors:__ [Jeremy Coyle](https://github.com/jeremyrcoyle) and [Nima
Hejazi](http://nimahejazi.org)

[![Travis-CI Build Status](https://travis-ci.org/jeremyrcoyle/mangolassi.svg?branch=master)](https://travis-ci.org/jeremyrcoyle/mangolassi)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/jeremyrcoyle/mangolassi?branch=master&svg=true)](https://ci.appveyor.com/project/jeremyrcoyle/mangolassi)
[![Coverage Status](https://img.shields.io/codecov/c/github/jeremyrcoyle/mangolassi/master.svg)](https://codecov.io/github/jeremyrcoyle/mangolassi?branch=master)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)

---

## Description

`hal9001` is an R package providing an implementation of the Highly Adaptive
LASSO (HAL), a nonparametric regression estimator that applies L1-regularized
regression (i.e., the LASSO) to a design matrix composed of indicator functions
corresponding to a set of covariates and interactions thereof, in a standard
statistical learning problem. Recent theoretical results show that HAL is
endowed with several important properties that make it optimally suited for
the purpose of inference in problem settings where causal parameters are
estimated via data-adaptive techniques (i.e., machine learning), as is the case
within the framework of Targeted Minimum Loss-Based Estimation (TMLE). While it
is certainly possible to implement HAL purely in R, the computationally
intensive nature of the algorithm suggests that writing core routines in C++
(and making these available in R via the [Rcpp](http://www.rcpp.org/) framework)
ought to provide significant efficiency gains. `hal9001` is just such an
implementation.

For detailed discussions of the Highly Adaptive LASSO estimator, the interested
reader might consider consulting @benkeser2016hal, @vdl2015generally,
and @vdl2017finite.

__This project is still in its infancy. Accordingly, it will be regularly
updated, possibly with breaking changes. For now, use only at your own risk.__

---

## Installation

<!--
For standard use, we recommend installing the package from
[CRAN](https://cran.r-project.org/) via

```{r cran-installation, eval = FALSE}
install.packages("hal9001")
```
-->

You can install the development version of `mangolassi` from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/) with

```{r gh-installation, eval = FALSE}
devtools::install_github("jeremyrcoyle/mangolassi")
```

---

## License

&copy; 2017 [Jeremy R. Coyle](https://github.com/jeremyrcoyle) & [Nima S.
Hejazi](http://nimahejazi.org)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

---

## References

