---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`lassi`

> A faster Highly Adaptive LASSO estimator

__Authors:__ [Jeremy Coyle](https://github.com/jeremyrcoyle) and [Nima
Hejazi](http://nimahejazi.org)

---

## Description

`lassi` is an R package providing utilities for performing a specialized form of
penalized regression (via application of the LASSO) as part of the Highly
Adaptive LASSO (HAL), a nonparametric regression estimator with theoretical
properties whose optimality has only recently been discovered. Due to the
computationally intensive nature of the HAL algorithm, core routines are written
in C++ and made available in R via the [Rcpp](http://www.rcpp.org/) framework.

__This project is still in its infancy. Accordingly, it will be regularly
updated, possibly with breaking changes. Use at your own risk.__

---

## Installation

<!--
For standard use, we recommend installing the package from
[CRAN](https://cran.r-project.org/) via

```{r cran-installation, eval = FALSE}
install.packages("lassi")
```
-->

You can install the development version of `lassi` from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/) with

```{r gh-installation, eval = FALSE}
devtools::install_github("nhejazi/lassi")
```

---

## License

&copy; 2017 [Jeremy R. Coyle](https://github.com/jeremyrcoyle) & [Nima S.
Hejazi](http://nimahejazi.org)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

