---
output:
  rmarkdown::github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`hal9001`

[![Travis-CI Build Status](https://travis-ci.org/tlverse/hal9001.svg?branch=master)](https://travis-ci.org/tlverse/hal9001)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/jeremyrcoyle/hal9001?branch=master&svg=true)](https://ci.appveyor.com/project/jeremyrcoyle/hal9001)
[![Coverage Status](https://img.shields.io/codecov/c/github/tlverse/hal9001/master.svg)](https://codecov.io/github/tlverse/hal9001?branch=master)
[![CRAN](http://www.r-pkg.org/badges/version/hal9001)](http://www.r-pkg.org/pkg/hal9001)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/hal9001)](https://CRAN.R-project.org/package=hal9001)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3558314.svg)](https://doi.org/10.5281/zenodo.3558314)

> The _Scalable_ Highly Adaptive Lasso

__Authors:__ [Jeremy Coyle](https://github.com/tlverse), [Nima
Hejazi](https://nimahejazi.org), and [Mark van der
Laan](https://vanderlaan-lab.org/)

---

## What's `hal9001`?

`hal9001` is an R package providing an implementation of the scalable _highly
adaptive lasso_ (HAL), a nonparametric regression estimator that applies
L1-regularized regression (i.e., the lasso) to a design matrix composed of
indicator functions corresponding to a set of covariates and interactions
thereof. Recent theoretical results show that HAL is endowed with several
important optimality properties, making it well-suited for the estimation of
highly complex functional forms while attaining fast convergence rates
($n^(1/4)$ and better) when used in the estimation of nuisance functions. HAL
has been quite successfully used in the construction of estimators at the
intersection of semiparametric theory and nonparametric causal inference (e.g.,
the construction of efficient one-step or targeted minimum loss estimators).

For detailed discussions of the highly adaptive lasso estimator, consider
consulting @benkeser2016hal, @vdl2017generally, and @vdl2017finite, among other
recent works.

---

## Installation

<!--
For standard use, we recommend installing the package from
[CRAN](https://cran.r-project.org/) via

```{r cran-installation, eval = FALSE}
install.packages("hal9001")
```
-->

To contribute, install the _development version_ of `hal9001` from GitHub via
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval = FALSE}
remotes::install_github("tlverse/hal9001", build_vignettes = FALSE)
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/tlverse/hal9001/issues).

---

## Example

This minimal example shows how to use `hal9001` to obtain predictions based on
the Highly Adaptive Lasso. For details on the properties of the estimator, the
interested reader is referred to @benkeser2016hal and @vdl2017generally.

```{r example}
# load the hal9001 package
library(hal9001)

# simulate data
set.seed(385971)
n <- 100
p <- 3
x <- xmat <- matrix(rnorm(n * p), n, p)
y <- x[, 1] * sin(x[, 2]) + rnorm(n, mean = 0, sd = 0.2)

# fit the HAL regression
hal_fit <- fit_hal(X = x, Y = y)
hal_fit$times

# training sample prediction
preds <- predict(hal_fit, new_data = x)
mean(hal_mse <- (preds - y)^2)
```

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/tlverse/hal9001/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `hal9001` R package, please cite the following:

        @manual{coyle2019hal9001,
          author = {Coyle, Jeremy R and Hejazi, Nima S and {van der Laan}, Mark
            J},
          title = {{hal9001}: The scalable highly adaptive lasso},
          year  = {2019},
          howpublished = {\url{https://github.com/tlverse/hal9001}},
          note = {{R} package version 0.2.5},
          url = {https://doi.org/10.5281/zenodo.3558314},
          doi = {10.5281/zenodo.3558314}
        }

---

## License

&copy; 2017-2020 [Jeremy R. Coyle](https://github.com/tlverse) & [Nima S.
Hejazi](https://nimahejazi.org)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

---

## References

