---
title: "Benchmarks for the `hal9001` Package"
author: "[Nima Hejazi](https://nimahejazi.org)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: vignette-refs.bib
vignette: >
  %\VignetteIndexEntry{Benchmarks for the `hal9001` Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, results='hide'}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
set.seed(7194568)

library(ggplot2)
library(wesanderson)
library(stringr)
library(tidyverse)

library(future)
library(microbenchmark)

library(SuperLearner)
library(hal9001)
```

## Introduction

This document consists of some simple benchmarks for various choices of the
`hal9001` implementation. The purpose of this document is two-fold:

1. Compare the computational performance of these methods
2. Illustrate the use of these different methods

```{r mse_fun}
# easily compute MSE
mse <- function(preds, y) {
  mean((preds - y)^2)
}
```

---

## The Continuous Outcome Case

### Standard use-cases with simple data structures

```{r sim}
# generate simple test data
n = 100
p = 3
x <- xmat <- matrix(rnorm(n * p), n, p)
y <- sin(x[, 1]) * sin(x[, 2]) + rnorm(n, 0.2)

testn = 10000
testx <- matrix(rnorm(testn * p), testn, p)
testy <- sin(testx[, 1]) * sin(testx[, 2]) + rnorm(testn, 0.2)
```

```{r hal_glmnet}
# glmnet implementation
glmnet_hal_fit <- fit_hal(X = x, Y = y, fit_type = "glmnet")
glmnet_hal_fit$times

# training sample prediction
preds_glmnet <- predict(glmnet_hal_fit, new_data = x)
glmnet_hal_mse <- mse(preds_glmnet, y)

# out-of-bag prediction
oob_preds_glmnet <- predict(glmnet_hal_fit, new_data = testx)
oob_glmnet_hal_mse <- mse(oob_preds_glmnet, y = testy)
```

```{r hal_origami}
# origami implementation
origami_hal_fit <- fit_hal(X = x, Y = y, fit_type = "origami")
origami_hal_fit$times

# training sample prediction
preds_origami <- predict(origami_hal_fit, new_data = x)
origami_hal_mse <- mse(preds_origami, y)

# out-of-bag prediction
oob_preds_origami <- predict(origami_hal_fit, new_data = testx)
oob_origami_hal_mse <- mse(oob_preds_origami, y = testy)
```

```{r hal_microbenchmark}
mb_hal_origami <- microbenchmark(unit = "s", times = 20,
  hal_fit_origami <- fit_hal(Y = y, X = x, fit_type = "origami")
)
summary(mb_hal_origami)

mb_hal_glmnet <- microbenchmark(unit = "s", times = 20,
  hal_fit_glmnet <- fit_hal(Y = y, X = x, fit_type = "glmnet")
)
summary(mb_hal_glmnet)

# visualize
p_fit_times <- rbind(mb_hal_origami, mb_hal_glmnet) %>%
  data.frame %>%
  transmute(
    type = ifelse(str_sub(as.character(expr), 9, 14) == "glmnet", "HAL-glmnet",
                  "HAL-origami"),
    time = time / 1e9
  ) %>%
  group_by(type) %>%
  ggplot(., aes(x = type, y = time, colour = type)) + geom_boxplot() +
    geom_point() + stat_boxplot(geom = 'errorbar') +
    scale_color_manual(values = wes_palette("GrandBudapest")) +
    xlab("") + ylab("time (sec.)") + ggtitle("") +
    theme_bw() + theme(legend.position = "none")
p_fit_times
```

### Advanced use-cases, with the `drtmle` R package

```{r sim_drtmle}
makeData <- function(n = n) {
    L0 <- data.frame(x.0 = runif(n, -1, 1))
    A0 <- rbinom(n, 1, plogis(L0$x.0 ^ 2))
    L1 <- data.frame(x.1 = L0$x.0 ^ 2 * A0 + runif(n))
    A1 <- rbinom(n, 1, plogis(L0$x.0 * L1$x.1))
    L2 <- rnorm(n, L0$x.0 ^ 2 * A0 * A1 + L1$x.1)
    return(list(L0 = L0, L1 = L1, L2 = L2, A0 = A0, A1 = A1))
}
dat <- makeData(n = 200)

# add drtmle defaults
abar <- c(1, 1)
stratify <- TRUE
```

```{r hal_benchmark_drtmle_estimateG}
mb_hal_origami_estimateG <- microbenchmark(unit = "s", times = 20,
  hal_origami_estimateG <- fit_hal(Y = as.numeric(dat$A0 == abar[1]),
                                   X = dat$L0,
                                   fit_type = "origami")
)
summary(mb_hal_origami_estimateG)

mb_hal_glmnet_estimateG <- microbenchmark(unit = "s", times = 20,
  hal_glmnet_estimateG <- fit_hal(Y = as.numeric(dat$A0 == abar[1]),
                                  X = dat$L0,
                                  fit_type = "glmnet")
)
summary(mb_hal_glmnet_estimateG)

# visualize
p_estimateG_times <- rbind(mb_hal_origami_estimateG,
                           mb_hal_glmnet_estimateG) %>%
  data.frame %>%
  transmute(
    type = ifelse(str_detect(as.character(expr), "glmnet"), "HAL-glmnet",
                  "HAL-origami"),
    time = time / 1e9
  ) %>%
  group_by(type) %>%
  ggplot(., aes(x = type, y = time, colour = type)) + geom_boxplot() +
    geom_point() + stat_boxplot(geom = 'errorbar') +
    scale_color_manual(values = wes_palette("GrandBudapest")) +
    xlab("") + ylab("time (sec.)") + ggtitle("") +
    theme_bw() + theme(legend.position = "none")
p_estimateG_times
```

```{r hal_benchmark_drtmle_estimateQ}
Y_estQ <- if (stratify) {
  dat$L2[dat$A0 == abar[1] & dat$A1 == abar[2]]
} else {
  dat$L2
}
X_estQ = as.numeric(unlist(ifelse(stratify,
                                  cbind(dat$L0, dat$L1)[dat$A0 == abar[1] &
                                                        dat$A1 == abar[2],],
                                  cbind(dat$L0, dat$L1, dat$A0, dat$A1))))

mb_hal_origami_estimateQ <- microbenchmark(unit = "s", times = 20,
  hal_origami_estimateQ <- fit_hal(Y = Y_estQ,
                                   X = X_estQ,
                                   fit_type = "origami")
)
summary(mb_hal_origami_estimateQ)


mb_hal_glmnet_estimateQ <- microbenchmark(unit = "s", times = 20,
  hal_glmnet_estimateQ <- fit_hal(Y = Y_estQ,
                                  X = X_estQ,
                                  fit_type = "glmnet")
)
summary(mb_hal_glmnet_estimateQ)

# visualize
p_estimateQ_times <- rbind(mb_hal_origami_estimateQ,
                           mb_hal_glmnet_estimateQ) %>%
  data.frame %>%
  transmute(
    type = ifelse(str_detect(as.character(expr), "glmnet"), "HAL-glmnet",
                  "HAL-origami"),
    time = time / 1e9
  ) %>%
  group_by(type) %>%
  ggplot(., aes(x = type, y = time, colour = type)) + geom_boxplot() +
    geom_point() + stat_boxplot(geom = 'errorbar') +
    scale_color_manual(values = wes_palette("GrandBudapest")) +
    xlab("") + ylab("time (sec.)") + ggtitle("") +
    theme_bw() + theme(legend.position = "none")
p_estimateQ_times
```

### Using `hal9001` with `SuperLearner`

The `hal9001` R package includes a learner wrapper for use with the
`SuperLearner` R package. In order to use this learner, the user need only add
"SL.hal9001" to their `SL.library` input. Here, we are interested in the
difference in the performance of `SuperLearner` when the `hal9001` wrapper uses
the `origami` backend versus the `glmnet` backend for the LASSO computation. To
facilitate this, we create two new SL wrappers from the one included:

```{r hal_sl_wrappers}
SL.hal9001.origami <- function (..., fit_type = "origami") {
  SL.hal9001(..., fit_type = fit_type)
}

SL.hal9001.glmnet <- function (..., fit_type = "glmnet") {
  SL.hal9001(..., fit_type = fit_type)
}
```


```{r hal_benchmark_drtmle_estimateG_SL}
mb_hal_origami_sl_estimateG <- microbenchmark(unit = "s", times = 20,
  hal_origami_sl_estimateG <- SuperLearner(Y = as.numeric(dat$A0 == abar[1]),
                                           X = dat$L0,
                                           SL.library = c("SL.speedglm",
                                                          "SL.hal9001.origami")
                                          )
)
summary(mb_hal_origami_sl_estimateG)

mb_hal_glmnet_sl_estimateG <- microbenchmark(unit = "s", times = 20,
  hal_glmnet_sl_estimateG <- SuperLearner(Y = as.numeric(dat$A0 == abar[1]),
                                          X = dat$L0,
                                          SL.library = c("SL.speedglm",
                                                         "SL.hal9001.glmnet")
                                         )
)
summary(mb_hal_glmnet_sl_estimateG)

# visualize
p_sl_estimateG_times <- rbind(mb_hal_origami_sl_estimateG,
                              mb_hal_glmnet_sl_estimateG) %>%
  data.frame %>%
  transmute(
    type = ifelse(str_detect(as.character(expr), "glmnet"), "HAL-glmnet",
                  "HAL-origami"),
    time = time / 1e9
  ) %>%
  group_by(type) %>%
  ggplot(., aes(x = type, y = time, colour = type)) + geom_boxplot() +
    geom_point() + stat_boxplot(geom = 'errorbar') +
    scale_color_manual(values = wes_palette("GrandBudapest")) +
    xlab("") + ylab("time (sec.)") + ggtitle("") +
    theme_bw() + theme(legend.position = "none")
p_sl_estimateG_times
```


<!--

## The Binary Outcome Case


```{r sim_binary}
# generate simple test data
n = 100
p = 3
x <- xmat <- matrix(rnorm(n * p), n, p)
y <- sin(x[, 1]) * sin(x[, 2]) + rnorm(n, 0.2)
y_bin <- as.numeric(y > median(y))
```

-->

