---
title: "Introduction to the HAL estimator"
author: "Nima Hejazi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: vignette-refs.bib
vignette: >
  %\VignetteIndexEntry{Introduction to the HAL estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The _highly adaptive LASSO_ (HAL) is a flexible machine learning algorithm that
nonparametrically estimates a function based on available data by embedding a
set of input observations and covariates in an extremely high-dimensional space
(i.e., generating basis functions from the available data). For an input data
matrix of $n$ observations and $d$ covariates, the number of basis functions
generated is approximately $n \cdot 2^{d - 1}$. To select a set of basis
functions from among the full set generated, the LASSO is employed. The
`hal9001` R package provides an efficient implementation of this routine,
relying on the `glmnet` R package for compatibility with the canonical LASSO
implementation while still providing a (faster) custom C++ routine for using the
LASSO with an input matrix composed of indicator functions. Consider consulting
the following references for more detailed descriptions of HAL:

1. [@benkeser2016hal],
2. [@vdl2015generally],
3. [@vdl2017finite].

---

## Preliminaries

```{r setup, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(microbenchmark)
```

```{r sim-data}
# simulation constants
set.seed(467392)
n_obs <- 1000
n_covars <- 3

# make some training data
x <- replicate(n_covars, rnorm(n_obs))
y <- sin(x[, 1]) * sin(x[, 2]) + rnorm(n_obs, mean = 0, sd = 0.2)

# make some testing data
test_x <- replicate(n_covars, rnorm(n_obs))
test_y <- sin(x[, 1]) * sin(x[, 2]) + rnorm(n_obs, mean = 0, sd = 0.2)
```

Look at simulated data...

```{r sim-view}
head(x)
head(y)
```

---

## Using the Highly Adaptive LASSO

```{r}
library(hal9001)
```

### Fitting the model

```{r fit-hal}
hal_fit <- fit_hal(X = x, Y = y, fit_type = "lassi")
hal_fit$times
```

```{r results-hal}
hal_fit
```

...

### Obtaining model predictions

```{r eval-mse}
# training sample prediction for HAL vs HAL9000
mse <- function(preds, y) {
    mean((preds - y)^2)
}

preds_hal <- predict(object = hal_fit, new_data = x)
mse_hal <- mse(preds = preds_hal, y = y)
mse_hal
```

```{r eval-oob}
oob_hal <- predict(object = hal_fit, new_data = test_x)
oob_hal_mse <- mse(preds = oob_hal, y = test_y)
oob_hal_mse
```

---

## References

